# Appendix

### Expected frequencies

Notice that R is telling us that the minimum expected frequency is 41.68. Why? For the Chi-squared test to work, it assumes the cell counts are sufficiently large. Precisely what constitutes 'sufficiently large' is a matter of some debate. One rule of thumb is that all expected cell counts should be above 5. If we have small cells, one alternative is to rely on the Fisher's Exact Test rather than on the Chi-Square. We don't have to request it here. Our cells are large enough for Chi Square to work fine. But if we needed, we could obtain the Fisher's Exact Test with the following code:

```{r, warning=FALSE, message=FALSE}
BCS0708<-read.csv("https://raw.githubusercontent.com/uom-resquant/modelling_book/refs/heads/master/datasets/BCS0708.csv")
library(gmodels)
BCS0708$rubbcomm <- as.factor(BCS0708$rubbcomm)
BCS0708$rubbcomm <- factor(BCS0708$rubbcomm, 
                    levels = c("not at all common", 
                    "not very common", "fairly common",
                    "very common"))
fisher.test(BCS0708$rubbcomm, BCS0708$bcsvictim, simulate.p.value=TRUE)
```

The p-value is still considerably lower than our alpha level of .05. So, we can still conclude that the relationship we observe can be generalised to the population. 

Remember that we didn't need the Fisher test. However, as suggested above, there may be times when you need them.

## Logistic regression

### Fitting logistic regression: alternative

Another way of fitting a logistic regression and getting the odds ratio with less typing is to use the `Logit()` function in the `lessR` package (you will need to install it if you do not have it).

```{r, warning=FALSE, message=FALSE}
library(effects)
data(Arrests, package="effects")

Arrests$harsher <- relevel(Arrests$released, "Yes")
#Rename the levels so that it is clear we now mean yes to harsher treatment
levels(Arrests$harsher) <- c("No","Yes")
#Check that it matches in reverse the original variable

Arrests$colour <- relevel(Arrests$colour, "White")
library(lessR, quietly= TRUE)
Logit(harsher ~ checks + colour + sex + employed, data=Arrests, brief=TRUE)
```
### Assessing model fit: deviance and pseudo r squared

As you may remember, when looking at linear models, we could use an F test to check the overall fit of the model, and we could evaluate R squared. When running logistic regression, we cannot obtain the R squared (although there is a collection of pseudo-R^2 measures that have been produced). In linear regression, things are a bit simpler. As Menard (2010: 43) explains:

> "there is only one reasonable residual variation criterion for quantitative variables in OLS, the familiar error sum of squares... but there are several possible residual variation criteria (entropy, squared error, qualitative difference) for binary variables. Another hindrance is the existence of numerous mathematical equivalents to R^2 in OLS, which are not necessarily mathematically (same formula) or conceptually (same meaning in the context of the model) equivalent to R^2 in logistic regression... Moreover, in logistic regression, we must choose whether we are more interested in qualitative prediction (whether predicitons are correct or incorrect), quantitative prediction (how close predictions are to being correct), or both, because different measures of explained variation are appropriate for these two different types of prediction"

A common starting point for assessing model fit is to look at the **log-likelihood statistic** and the **deviance** (also referred to as -2LL). 

The log-likelihood aims to provide a measure of how much-unexplained variation there is after you fit the mode. Large values indicate poor fit. 

The deviance, on the other hand, is simply the log-likelihood multiplied by -2 and is generally abbreviated as -2LL. The deviance will be a positive value, and *larger values indicate worse prediction* of the response variable. It is analogous to the error sum of squares in linear regression. In the same way that OLS linear regression tries to minimise the error sum of squares, maximum likelihood logistic regression tries to minimise the -2LL.

The difference between the -2LL for the model with no predictors and the -2LL for the model with all the predictors is the closer we get in logistic regression to the regression sum of squares. This difference is often called **model chi-squared**, and it provides a test of the null hypothesis that all the regression coefficients equal zero. It is, thus, equivalent to the F test in OLS regression.

```{r, warning=FALSE, message=FALSE}
fitl_1 <- glm(harsher ~ checks + colour + sex + employed, data=Arrests, family = "binomial")
summary(fitl_1)

```

In our example, we saw that some measures of fit were printed below the table with the coefficients. The **null deviance** is the deviance of the model with no predictors, and the **residual deviance** is simply the deviance of this model. You clearly want the residual deviance to be smaller than the null deviance. The difference between the null and the residual deviance is what we call the model chi-squared. In this case, this is 4776.3 minus 4330.7. We can ask R to do this for us.

First, notice that the object we created has all the information we need already stored.

```{r}
names(fitl_1)
```

So we can use this stored information in our calculations.

```{r}
with(fitl_1, null.deviance - deviance)
```

Is `r 4776.3-4330.7` small? How much smaller is enough? This value has a chi-square distribution, and its significance can be easily computed. For this computation, we need to know the degrees of freedom for the model (which equal the number of predictors in the model) and can be obtained like this:

```{r}
with(fitl_1, df.null - df.residual)
```

Finally, the p-value can be obtained using the following code to invoke the Chi-Square distribution:

```{r}
#When doing it yourself, this is all you really need 
#(we present the code in a separate fashion above so that you understand better what the one here does)
with(fitl_1, pchisq(null.deviance - deviance, df.null - df.residual, lower.tail = FALSE))
```

We can see that the model chi-square is highly significant. Our model as a whole fits significantly better than a model with no predictors.

Menard (2010) recommends also looking at the likelihood ratio R^2, which can be calculated as the difference between the null deviance and the residual deviance divided by the null deviance.

```{r}
#Likelihood ratio R2
with(fitl_1, (null.deviance - deviance)/null.deviance)
```

Some authors refer to this as the Hosmer/Lemeshow R^2. It indicates how much the inclusion of the independent variables in the model reduces variation, as measured by the null deviance. It varies between 0 (when our prediction is catastrophically useless) and 1 (when we predict with total accuracy). There are many other pseudo R^2 measures that have been proposed, but Menard [based on research](http://www.tandfonline.com/doi/pdf/10.1080/00031305.2000.10474502) on the properties of various of these measures recommends the likelihood ratio R^2 because:

+ It is the one with a closer conceptual link to R^2 in OLS regression.

+ It does not appear to be sensitive to the base rate (the proportion of cases that have the attribute of interest) of the phenomenon being studied and, therefore, will work even in cases with unbalanced probabilities.

+ It varies between 0 and 1

+ And it can be used in other generalised linear models (models for categorical outcomes with more than two levels, which we don't cover here)

### Assessing model fit: ROC curves

We may want to see what happens to sensitivity and specificity for different cut-off points. For this, we can look at **receiver operating characteristics** or simply [ROC curves](http://en.wikipedia.org/wiki/Receiver_operating_characteristic). This is essentially a tool for evaluating the sensitivity/specificity trade-off. The ROC curve can be used to investigate alternate cut-offs for class probabilities.

We can use the `pROC` package for this. We start by creating an object that contains the relevant information with the `roc()` function from the `pROC` package.

```{r, warning=FALSE, message=FALSE}

fitl_1_prob <- predict(fitl_1, type = "response") 
library(pROC)
rocCURVE <- roc(response = Arrests$harsher, 
                predictor = fitl_1_prob)
```

Once we have the object with the information, we can plot the ROC curve.

```{r}
plot(rocCURVE, legacy.axes = TRUE) #By default, the x-axis goes backwards; we can use the specified option legacy.axes=TRUE, to get 1-spec on the x-axis moving from 0 to 1.
```

We can see the trajectory of the curve is at first steep, suggesting that sensitivity increases at a greater pace than the decrease in specificity. However, we then reach a point at which specificity decreases at a greater rate than the sensitivity increases. If you want to select a cut-off that gives you the optimal cut-off point, you can use the `coords()` function of the pROC package. You can pass arguments to this function so that it returns the best sum of sensitivity and specificity.

```{r}
alt_cutoff1 <- coords(rocCURVE, x = "best", best.method = "closest.topleft")
#The x argument, in this case, is selecting the best cut-off using the "closest topleft" method 
#(which identifies the point closest to the top-left part of the plot with perfect sensitivity or specificity). Another option is to use the "youden" method in the best.method argument.
alt_cutoff1
```

Here, we can see that with a cut-off point of .16 we get a specificity of .63 and a sensitivity of .71. Notice how this is close to the base rate of harsher treatment in the sample (17% of individuals actually received harsher treatment). For a more informed discussion of cut-off points and costs of errors in applied predictive problems in criminal justice, I recommend reading [Berk (2012)](http://link.springer.com/book/10.1007%2F978-1-4614-3085-8). Often, the selection of cut-off may be motivated by practical considerations (e.g., selecting individuals for treatment in a situation where resources to do so are limited).

The ROC curve can also be used to develop a quantitative assessment of the model. The perfect model is one where the curve reaches the top left corner of the plot. This would imply 100% sensitivity and specificity. On the other hand, a useless model would be one with a curve alongside the diagonal line splitting the plot in two, from the bottom right corner to the top right corner. You can also look at the **area under the curve** (AUC) and use it to compare models. An AUC of .5 corresponds to the situation where our predictors have no predictive utility. For a fuller discussion of how to compare these curves and the AUC, I recommend reading Chapter 11 of [Kuhn and Johnson (2014)](http://link.springer.com/book/10.1007/978-1-4614-6849-3).





